---
title: "3: Data Exploration"
author: "Environmental Data Analytics | John Fay & Luana Lima"
date: "Fall 2025"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Import and explore datasets in R
2. Graphically explore datasets in R
3. Apply data exploration skills to a real-world example dataset

---

## Discussion: Why do we explore our data?

Why is data exploration our first step in analyzing a dataset? What information do we gain? How does data exploration aid in our decision-making for data analysis steps further down the pipeline?

>Answer: It's important to know they type and availability of data in a dataset. In particular, which columns are optimal, types of units, missing units, etc. It allows us to make informed decisions as we continue processing because we know what we have available and whether it fits our needs.

---

## Import data and view summaries

```{r, Setup-Import-Explore, message = FALSE}
## -Set up-
# 1. Load packages (tidyverse, here)
library(tidyverse)
library(here)

# 2. Check the working directory
getwd()
here()

# 3. Import datasets
USGS.flow.data <- read.csv(
  here("Data/Processed/USGS_Site02085000_Flow_Processed.csv"),
  stringsAsFactors = TRUE,
  colClasses = c('site_no' = 'factor')
)

## -Initial Exploration-

#View(USGS.flow.data)
# Alternate option: click on data frame in Environment tab

#View column names and the structure of the dataset
colnames(USGS.flow.data)
str(USGS.flow.data)
# Check our date column: is it a date object?
class(USGS.flow.data$datetime)

# Change it to a date object and confirm
USGS.flow.data$datetime <- as.Date(USGS.flow.data$datetime, format = '%Y-%m-%d')
class(USGS.flow.data$datetime)

```

---

## Visualization for Data Exploration
Although the `summary()` function is helpful in getting an idea of the spread of values in a numeric dataset, it can be useful to create visual representations of the data to help form hypotheses and direct downstream data analysis. Below is a summary of the useful types of graphs for data exploration.

Note: each of these approaches utilize the package "ggplot2". We will be covering the syntax of ggplot in a later lesson, but for now you should familiarize yourself with the functionality of what each command is doing.

### Bar Chart (function: geom_bar)

Visualize count data for categorical variables. 

```{r, fig.height = 3, fig.width = 4}
ggplot(data = USGS.flow.data, aes(x = discharge.mean.approval)) +
  geom_bar()
```

### Histogram (function: geom_histogram)
Visualize distributions of values for continuous numerical variables. What is happening in each line of code? Insert a comment above each line.

```{r, fig.height = 3, fig.width = 4}
# it is plotting a frequency distribution of the mean discharge, but is not plotting the NA values
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean))

# Plotting a frequency distribution, but sets a bin width of 10 units, and does not plot NA values
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), binwidth = 10)

# Plots a frequency distribution with a set number of bins (20) and does not plot NA values
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), bins = 20)

# plots a frequency distribution with a bind with of 10 units, no NA values, and a set x-axis of 0 to 500
ggplot(USGS.flow.data, aes(x = discharge.mean)) +
  geom_histogram(binwidth = 10) + 
  scale_x_continuous(limits = c(0, 500))
  
# Plots a frequency distribution of mean gage height
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = gage.height.mean))

```


### Frequency line graph (function: geom_freqpoly)
An alternate to a histogram is a frequency polygon graph (distributions of values for continuous numerical variables). Instead of displaying bars,  counts of continuous variables are displayed as lines. This is advantageous if you want to display multiple variables or categories of variables at once.

```{r, fig.height = 3, fig.width = 4}
#Plots a frequency line for the min, mean, and max, with a set number of bins and varying symbology to differentiate the three fields. The x-axis is also limited at 0 to 10
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean), bins = 50) +
  geom_freqpoly(aes(x = gage.height.min), bins = 50, color = "darkgray") +
  geom_freqpoly(aes(x = gage.height.max), bins = 50,  lty = 2) +
  scale_x_continuous(limits = c(0, 10))

#frequency line of gage mean height and height mean approval, set to 50 bins on a x scale of 0 to 10. This colors the mean values by approval status and adds a legend at the top.
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean, color = gage.height.mean.approval), bins = 50) +
  scale_x_continuous(limits = c(0, 10)) +
  theme(legend.position = "top")

```


### Box-and-whisker plots (function: geom_boxplot, geom_violin)
A box-and-whisker plot is yet another alternative to histograms (distributions of values for continuous numerical variables). These plots consist of: 

* A box from the 25th to the 75th percentile of the data, called the interquartile range (IQR).

* A bold line inside the box representing the median value of the data. Whether the median is in the center or off to one side of the IQR will give you an idea about the skewness of your data.

* A line outside of the box representing values falling within 1.5 times the IQR. 

* Points representing outliers, values that fall outside 1.5 times the IQR. 

An alternate option is a violin plot, which displays density distributions, somewhat like a hybrid of the box-and-whiskers and the frequency polygon plot.

```{r, fig.height = 3, fig.width = 4}
# creates a box and whisker plot with height mean approval on the x-axis and gage mean height on the y
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean.approval, y = gage.height.mean))

#plots the mean gage height by mean discharge and groups them 
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean, y = discharge.mean, group = cut_width(gage.height.mean, 1)))

#creates a violin plot of height mean approval and mean height with added quartile lines.
ggplot(USGS.flow.data) +
  geom_violin(aes(x = gage.height.mean.approval, y = gage.height.mean), 
              draw_quantiles = c(0.25, 0.5, 0.75))
```

### Scatterplot (function: geom_point)
Visualize relationships between continuous numerical variables.

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data) +
  geom_point(aes(x = discharge.mean, y = gage.height.mean))

ggplot(USGS.flow.data) +
  geom_point(aes(x = datetime, y = discharge.mean))
```

Question: under what circumstances would it be beneficial to use each of these graph types (bar plot, histogram, frequency polygon, box-and whisker, violin, scatterplot)?

> Answer: The bar plot is good for looking at values plotted against categories, histograms are best for looking at frequency in different categories, frquency polygons can be used for plotting frequencies of multiple fields together, box-and-whisker plots work best for plotting data spread over quartiles, violin plots also are good for plotting the spread of data, and scatterplots are best for plotting two variables against each other. 

## Ending discussion
What did you learn about the USGS discharge dataset today? What separate insights did the different graph types offer?
> Answer: I learned that the data covers a significant time period. Because each graph type allows us to look at either different aspects of the data, or similar aspects in different visuals, we are able to see what types of relationships might exist, such as mean gage hight and discahrge which has a noticable relationship, versus gage height approval and mean gage height.


How can multiple options for data exploration inform our understanding of our data?

> Answer: It allows us to look at multiple aspects of the data in ways that are appropriate for multiple circumstance (for example, continuous data vs. categorical data). 


Do you see any patterns in the USGS data for the Eno River? What might be responsible for those patterns and/or relationships?

> Answer: There is a strong correlation between high mean gage height and mean discharge. This relationship is likely because the amount of discharge will affect the height of the water on the gage.


