---
title: "Assignment 5: Data Visualization"
author: "Sloane Hauck"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse);library(here);library(cowplot);library(ggplot2);library(ggthemes)

here()

Lake_Chemistry <- read.csv(
  file = here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE
)

Litter_Mass <- read.csv(
  file = here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"),
  stringsAsFactors = TRUE
)

#2 
class(Lake_Chemistry$sampledate)
Lake_Chemistry$sampledate <- ymd(Lake_Chemistry$sampledate)

class(Litter_Mass$collectDate)
Litter_Mass$collectDate <- ymd(Litter_Mass$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
my_theme <- theme_base()  + 
  theme(
    plot.title = element_text(size = 12,
                              color = "darkblue"),
    axis.title.x = element_text(size = 10,
                                color = "darkblue"),
    axis.title.y =  element_text(size = 10,
                                 color = "darkblue"),
    axis.ticks =  element_line(color = "black"),
    plot.background = element_rect(color = "black",
                                   fill = "lightgrey"),
    complete = TRUE
    )
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
ggplot(Lake_Chemistry, aes(x=po4, y=tp_ug, color = lakename)) +
  geom_point() +
  facet_wrap(facets = vars(lakename),
             nrow = 2) +
  ylab("Total Phosphorous") +
  xlab("Phosphate") +
  xlim(0,50) +
  labs(color = "Lake Name" ) +
  geom_smooth(method = lm, color = "navy") + 
  theme(my_theme)
 

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
Temp <- Lake_Chemistry %>% 
  ggplot(aes(x = factor(month, levels = 1:12, labels = month.abb), 
             y = temperature_C, fill = lakename)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
#  scale_x_discrete(drop = FALSE) +
  xlab("Month") +
  ylab("Temperature (C)") + 
  theme(my_theme, legend.position = "None")

Total_Phosphorous <- Lake_Chemistry %>% 
  ggplot(aes(x = factor(month, levels = 1:12, labels = month.abb), 
             y = tp_ug, fill = lakename)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
 # scale_x_discrete(drop = FALSE) +
  xlab("Month") +
  ylab("Total Phosphorous (ug)") +
  theme(my_theme, legend.position = "None")

Total_Nitrogen <- Lake_Chemistry %>% 
  ggplot(aes(x = factor(month, levels = 1:12, labels = month.abb), 
             y = tn_ug, fill = lakename)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
 # scale_x_discrete(drop = FALSE) +
  xlab("Month") +
  ylab("Total Nitrogen (ug)") +
  theme(my_theme) +
  labs(fill = "Lake Name")

plot_grid(Temp, Total_Nitrogen, Total_Phosphorous, nrow = 3,
          align = 'h', rel_heights = c(1, 1))
  

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature is higher in the summer months, as would be expected. Additionally, the nutrient levels are also slightly higher by the end of the summer/early autumn.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
Needles <- Litter_Mass %>% 
  filter(functionalGroup == "Needles") %>% 
  ggplot(aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_line() +
  theme(my_theme, axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_x_date(date_breaks = "3 month", date_labels = "%m-%Y") +
  xlab("Collection Date") +
  ylab("Dry Mass") +
  labs(color = "NLCD Class")

Needles

#7

Needles_Separate <- Litter_Mass %>% 
  filter(functionalGroup == "Needles") %>% 
  mutate(nlcdClass = case_when(
    nlcdClass %in% "evergreenForest" ~ "Evergreen - Forest",
    nlcdClass %in% "grasslandHerbaceous" ~ "Grassland - Herbaceous",
    nlcdClass %in% "shrubScrub" ~ "Shrub - Scrub",
    TRUE ~ nlcdClass
  )) %>% 
  ggplot(aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_line() +
  facet_wrap(vars(nlcdClass)) +
  theme(my_theme, axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "None") + 
  scale_x_date(date_breaks = "3 month", date_labels = "%m-%Y") +
  xlab("Collection Date") +
  ylab("Dry Mass") +
  labs(color = "NLCD Class")

Needles_Separate

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think that 7 is more effective because it is easier to see trends over time and by having the graphs sit next to each other, you can still see what overlapping trends exist.