---
title: "Lab 3a - Exercises"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Fall 2025"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Set up the session
- Load packaged: tidyverse, lubridate, & here
- Import the `EPAair_PM25_NC2019_raw.csv` file
  - Set strings to be read in as factors
  - Set the "Site.ID" column to be read in as a factor
- Set the date column to be a date

```{r Setup}
#Import packages
library(tidyverse); library(lubridate); library(here)

#Check workspace
here()

#Import data
pm25 <- read.csv(
  file = here('Data', 'Raw', 'EPAair_PM25_NC2019_raw.csv'),
  stringsAsFactors = TRUE,
  colClasses = c("Site.ID"='factor')
)

#Fix the date
pm25$Date <- mdy(pm25$Date)

```


## Explore properties of the dataframe
-   How many different sites are included in the data set?
-   What is the mean daily PM2.5 concentration?
-   In what units of are PM 2.5 concentrations reported?

```{r Explore the data}
# Compute the number of different sites
length(levels(pm25$Site.ID))

# Compute the mean and median PM 2.5 concentration
mean_pm25 <- mean(pm25$Daily.Mean.PM2.5.Concentration)
mean_pm25

median_pm25 <- median(pm25$Daily.Mean.PM2.5.Concentration)
median_pm25

# What are the units
unique(pm25$UNITS)

# What is the date range (min and max dates)
date_min <- min(pm25$Date)
date_min
date_max <- max(pm25$Date)
date_max

```


## Plot distribution of values

-   Number of samples using each different source 
-   Distribution of daily mean concentration values
-   Distribution of daily mean concentration, by source*
  *   There are multiple ways to visualize this...

```{r Visualize the data}
#Number of samples using each different source 
ggplot(pm25, aes(x=Source)) +
  geom_bar()


#Distribution of daily mean concentration values
ggplot(pm25, aes(x=Daily.Mean.PM2.5.Concentration)) +
  geom_histogram()


#Distribution of daily mean concentration, by source
ggplot(pm25, aes(x=Source, y=Daily.Mean.PM2.5.Concentration)) +
  geom_boxplot()



```

